NumPy---Array Programming in Python

### 2020-01-10 body.tex
Minor edit
Another round of edits on the array section

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy\cite{virtanen2019scipy},
matplotlib\cite{hunter2007matplotlib}, pandas\cite{mckinney-proc-scipy-2010},
scikit-learn\cite{pedregosa2011scikit}, and
scikit-image\cite{vanderwalt2014scikit}.
It consists of an $n$-dimensional array object along with utility functions
that operate on it.
Because of its inherent simplicity---being a pointer to memory with some
associated meta-data---the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for array programming.

@@

The NumPy array is a structure that efficiently stores and accesses
$n$-dimensional arrays, also known as tensors\cite{vanderwalt2011numpy}.
The need for such a data structure arises often in science, as it can be used to represent everything from time-series measurements to time-varying spectral measurements on a grid.

The NumPy array consists of a pointer to memory, along with meta-data used to
interpret the data stored there, notably {\em data type}, {\em shape}, and {\em strides} (see Figure~\ref{fig:array-concepts}).

@@

The {\em shape} of an array determines the number of axes, or dimensionality, of the array.  For example, a vector of numbers can be stored as a 1-D array of shape $N$, while color videos are 4-D arrays of shape $T \times M \times N \times 3$.

% shape > num. of axes and number of elements in each axis

@@

% on top of this structure we add a powerful, expressive syntax for indexing and element-wise
% calculations ....
% this high-level syntax allows us to be efficient (broadcasting) and implement in C for
% optimizing

Portions of an array can be accessed through an {\em indexing} operation.  Arrays can be indexed in several ways to return single elements, sub-arrays, or elements that satisfy a specific condition.  Arrays can even be indexed using other arrays (see Figure~\ref{fig:array-concepts}).  Wherever possible, indexing that retrieves a subarray returns a {\em view} on the original array, such that data is shared between the two arrays.  This provides a powerful way to operate on subsets of array data, while limiting memory usage.

Finally, a powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}. When performing an operation (such as addition) on two arrays, one intuitively expect that these arrays should be required to have the same shape.  However, through broadcasting, NumPy allows the dimensions to differ, while still producing results that appeal to intuition.  A trivial example is the addition of a scalar value to an array, but it also generalizes to more complex examples such as scaling each column of an array, or generating a grid of coordinates.  In broadcasting, one or both arrays are virtually duplicated (that is, without copying any data in memory), so that the shapes of the operands match (see Figure~\ref{fig:array-concepts}).  Broadcasting is also applied whenever an array is indexed using arrays of indices.% (see Figure~\ref{fig:broadcasting}).

Once data is viewed as an $n$-dimensional array, more abstract operations such
as reductions (e.g., summation across columns, see Figure~\ref{fig:reductions}) products, and so forth, are
defined.

@@

NumPy includes numerous functions that perform element-wise and reduction operations on arrays, including arithmetic, statistics, and trigonometry.  They aim to loop over array elements near-optimally, taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory.  Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.

@@

Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and, for matrix multiplication, $@$), together with {\em utility functions}; together, these provide an easily readable, expressive, high-level API (Application Programming Interface) for array manipulation.

@@

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.

For example, a new array protocol was implemented that allows NumPy to consume
array-like objects produced by libraries outside of NumPy.  When asking NumPy
to, e.g., sum such an array, instead of attempting that operation itself, NumPy
requests the original array to complete the computation.  This means that the
NumPy API---the syntax for interacting with arrays---can still be used, but now
with a library that supports GPUs, does distributed computation processing, or
stores data in the cloud.

Another ongoing example is a redesign of the data type system.
While the majority of users rely only on numerical dtypes, there are a
multitude of use cases for which the current implementation is unsuited, such
as physical units\cite{astropy,Goldbaum2018,pint}, geometrical
objects\cite{pygeos}, and automatic differentiation\cite{pyadolc}.
The proposed overhaul of the datatype system will make it behave consistently
and will simplify the creation of custom dtypes, both in C and in Python.

@@

The third factor is much like the second: 
%a culture in Python that concentrates
%on high quality of code and distribution.  Because Python is open-source, it has
%a strong culture of contribution, and therefore, of using tools and process to
%improve collaborative software development, such as distributed version
%control, code review, and automated testing.  These have allowed the project to
%scale safely as it attracted more users and developers, even though these
%developers rarely came to the project with training in software engineering.
the project has strong culture of using software-engineering practice to
improve collaboration and reduce error \cite{millman2014developing}.
The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy.
The project has comprehensive, high-quality documentation,
integrated with the source
code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

>>>>

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy \cite{virtanen2019scipy},
matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
scikit-learn \cite{pedregosa2011scikit}, and
scikit-image \cite{vanderwalt2014scikit}.
It consists of an $n$-dimensional array object along with utility functions
that operate on it.
Because of its inherent simplicity---being a pointer to memory with some
associated meta-data---the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for array programming.

@@

The NumPy array is a structure that efficiently stores and accesses $n$-dimensional arrays, also known as tensors \cite{vanderwalt2011numpy}, that enables a wide variety of scientific computation.  It consists of a pointer to memory, along with meta-data used to interpret the data stored there, notably a {\em data type}, {\em shape}, and {\em strides} (see Figure~\ref{fig:array-concepts}).

@@

The {\em shape} of an array determines the number of elements along each axis, and the number of axes is the array's dimensionality. For example, a vector of numbers can be stored as a one dimensional array of shape $N$, while color videos are four dimensional arrays of shape $T \times M \times N \times 3$.

@@

To manipulate the data structure in an intuitive manner, we provide users with a simple yet powerful array expression syntax.  It supports operations such as indexing (selecting subarrays) and element-wise calculations.  A high level syntax allows users to express themselves succinctly, while NumPy deals with the underlying mechanics of making operations fast.

Indexing an array returns single elements, sub-arrays, or elements that satisfy a specific condition.  Arrays can even be indexed using other arrays (see Figure~\ref{fig:array-concepts}).  Wherever possible, indexing that retrieves a subarray returns a {\em view} on the original array, such that data is shared between the two arrays.  This provides a powerful way to operate on subsets of array data, while limiting memory usage.

A powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}. When performing an operation (such as addition) on two arrays, one intuitively expect that these arrays should be required to have the same shape.  However, through broadcasting, NumPy allows the dimensions to differ, while still producing results that appeal to intuition.  A trivial example is the addition of a scalar value to an array, but it also generalizes to more complex examples such as scaling each column of an array, or generating a grid of coordinates.  In broadcasting, one or both arrays are virtually duplicated (that is, without copying any data in memory), so that the shapes of the operands match (see Figure~\ref{fig:array-concepts}).  Broadcasting is also applied whenever an array is indexed using arrays of indices.% (see Figure~\ref{fig:broadcasting}).

@@

%Once data is viewed as an $n$-dimensional array, more abstract operations such
%as reductions can be defined.

%(e.g., summation across columns, see Figure~\ref{fig:reductions}) products, and %so forth, are
%defined.

To complement the array syntax, NumPy includes several functions that perform element-wise calculations on arrays, including arithmetic, statistics, and trigonometry.  They aim to loop over array elements near-optimally, taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory.  Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.

Many of these functions, such as summation, support reductions: aggregating results across over or more dimension of the array.  For example, summing a $D$-dimensional array over $p$ axes results in a $D-p$-dimensional array, as shown in Figure~\ref{fig:reductions}.

@@

Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and, for matrix multiplication, $@$), as well as {\em utility functions}; together, these provide an easily readable, expressive, high-level API for array programming.  Exposing these array programming primitives in Python---an interpreted language---creates an ideal environment for exploratory data analysis \cite{perez2007ipython} where users may inspect, manipulate, and visualize their data.

\fixme{This end of paragraph needs fixing up}Instead of focusing on programming, users can bring their attention to understanding and interacting with their data.

@@

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
For example, a new array protocol was implemented that allows NumPy to consume
array-like objects produced by libraries outside of NumPy.  When asking NumPy
to, e.g., sum such an array, instead of attempting that operation itself, NumPy
requests the original array to complete the computation.  This means that the
NumPy API---the syntax for interacting with arrays---can still be used, but now
with a library that supports GPUs, does distributed computation processing, or
stores data in the cloud.

NumPy also aims to improve our \emph{extensibility}.
For example, we are currently redesigning the data type system.
While the majority of users rely only on numerical dtypes, there are a
multitude of use cases for which the current implementation is unsuited, such
as physical units \cite{astropy,Goldbaum2018,pint}, geometrical
objects \cite{pygeos}, and automatic differentiation \cite{pyadolc}.
The proposed overhaul of the datatype system will make it behave consistently
and will simplify the creation of custom dtypes, both in C and in Python.

@@

The third factor is much like the second: 
%a culture in Python that concentrates
%on high quality of code and distribution.  Because Python is open-source, it has
%a strong culture of contribution, and therefore, of using tools and process to
%improve collaborative software development, such as distributed version
%control, code review, and automated testing.  These have allowed the project to
%scale safely as it attracted more users and developers, even though these
%developers rarely came to the project with training in software engineering.
the project has strong culture of using software-engineering practice to
improve collaboration and reduce error \cite{millman2014developing}.
The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy.
The project has comprehensive, high-quality documentation,
integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
